---
title: "Cyclistic Case of Study"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Clear Statement of the Business Task

Analyze Cyclistic’s historical bike trip data to determine how annual members and casual riders use bikes differently. This includes evaluating ride patterns such as trip duration, times of usage, and station preferences. These insights will help the marketing team design a data-driven strategy to convert casual riders into annual members.

## Data Sources Description

The data for this analysis was obtained from Cyclistic’s historical bike trip records, provided by Motivate International Inc. The dataset contains a year's worth of anonymized information on individual bike trips taken within Cyclistic's bike-share program in Chicago. This data is essential for understanding user behavior, particularly the differences between casual riders and annual members, to inform future marketing strategies.

**Data Sources and Location**

- **Data Source**:

Historical trip data was publicly released by Motivate International Inc. and made available for analysis. The dataset is anonymized, ensuring no personally identifiable information (PII) is included.

- **Storage Location**:

The data was downloaded and stored in the year folder under the project directory. Each file represents a specific month, organized in CSV format.

### Key Variables and Structure

- **Rows**: Each row represents an individual bike trip.

- **Columns**: The dataset includes 13 variables, categorized as follows:

- **Trip Information**:

ride_id: Unique identifier for each trip. 
rideable_type: Type of bike used (e.g., standard bike, cargo bike). started_at and ended_at: Timestamps for trip start and end. start_station_name and end_station_name: Names of starting and ending stations.

- **Geographical Information**: 
start_lat, start_lng: Latitude and longitude of the starting station. end_lat, end_lng: Latitude and longitude of the ending station.

- **Rider Type**: 
member_casual: Indicates if the rider is a casual user or an annual member.

### Considerations for Data Integrity and Credibility

The dataset adheres to ROCCC principles: 

Reliability: Provided by a trusted source (Motivate International Inc.). 
Originality: Represents real-world usage data. 
Comprehensiveness: Covers a full year, capturing seasonal trends. 
Consistency: Uniform schema across all monthly files. 
Currency: While historical, the data is recent enough to provide actionable insights. 

Bias Considerations: 
Limited demographic data restricts deeper segmentation. 
Potential overrepresentation of frequent riders or urban commuters. 

**Licensing and Privacy**

Data is publicly available under a license provided by Motivate International Inc. 
Personally Identifiable Information has been excluded to ensure compliance with privacy standards.

### Data Cleaning and Preparation 

-**Data Combination**: Monthly files were combined into a single data frame (all_trip_data) using the bind_rows() function for a comprehensive analysis. 

Despite these limitations, the dataset provides valuable insights into Cyclistic users' riding patterns and can support data-driven marketing strategies.

-**Data Cleaning Steps**: 

1. **Data Validation**

  1.1 Column data types (chr, dbl, dttm) were verified using glimpse() to ensure compatibility with the analysis.
  1.2 A total of 1,298,134 rows with missing values were identified using is.na() and colSums(is.na()).
  1.3 Missing values in key columns were handled by replacing NA values in station names with "Unknown Start Station" and "Unknown   1.4 End Station", and by using the median for missing coordinates.
0 duplicate rows were detected and removed, ensuring dataset uniqueness.
  1.5 Date and time formats (started_at, ended_at) were validated to ensure correct parsing.

2. **Transforming the Data**

  2.1 Ride Duration Calculation: A new column, ride_length, was created by calculating the difference between ended_at and started_at in minutes using difftime().

  2.2 Filtering Out Invalid Rides: Records where ride_length was less than or equal to 0 were removed to ensure data validity.

  2.3 Day of the Week Calculation: A new column, day_of_week, was created using wday() to analyze usage trends by the day of the week.

  2.4 Ensuring Consistent Data Formats: Validated and corrected data types (e.g., timestamps were parsed to dttm). 
Ensured ride durations were positive by calculating ride_length and filtering out negative values.

## Analysis of Ride Data: Calculations, Trends, and Relationships

1. **Performing Calculations**

- Ride Duration: The average ride duration is significantly higher for casual users (~29 minutes) compared to members (~13 minutes).
- Most Frequent Ride Day: Saturday is the most common day for rides.
- Maximum Ride Length: The longest recorded ride lasted 41,387.25 minutes (~28 days), likely an outlier.
- Most used stations: It is difficult to draw definitive conclusions due to inconsistent data. As mentioned earlier, there were missing values in the dataset. Instead of deleting those rows, I decided to fill them as "Unknown Start Station" to preserve as much data as possible.

2. **Identifying Trends and Relationships**

- User Behavior Differences: 
Casual users tend to have longer ride durations than members, suggesting they may use bikes for leisure rather than commuting.
- Weekly Trends: 
Members take more rides on weekdays, particularly Tuesday through Thursday, while casual users ride more on weekends, with a peak on Saturday.
- Hourly Trends: 
Members' rides peak during commuting hours (7-9 AM and 5-7 PM), indicating work-related use.
Casual users ride more steadily throughout the day, with a small evening peak.
- Seasonal Patterns:
Rides increase from spring, peak in summer (July and August), and decline in fall and winter.
Casual users' ridership drops significantly in colder months, while members maintain a relatively stable pattern.


This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

```{r}
# Install necessary packages
library(tidyverse)
library(lubridate)
library(readxl)
library(conflicted)

# Define the folder path containing CSV files
folder_path <- "/Users/viviana/Documents/CYCLISTIC/year"
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)

# Read all CSV files and store them in a list
trip_data_list <- lapply(csv_files, read_csv)

# Combine all datasets into one
all_trip_data <- bind_rows(trip_data_list)

# Check the dimensions and first rows of the combined data
dim(all_trip_data)
head(all_trip_data)

# Check the structure of the combined dataset
glimpse(all_trip_data)

# Summary statistics for numerical columns
summary(all_trip_data)

# Create new columns for ride length and day of the week
all_trip_data <- all_trip_data %>%
  mutate(
    ride_length = as.numeric(difftime(ended_at, started_at, units = "mins")),
    day_of_week = wday(started_at, label = TRUE)
  ) %>%
  dplyr::filter(ride_length > 0)  # Remove invalid data

# Remove rows with missing or inconsistent data
all_trip_data <- drop_na(all_trip_data)

# Identify duplicate rows
duplicated_rows <- all_trip_data %>%
  dplyr::filter(duplicated(.))

# Print duplicate rows
print(duplicated_rows)

# Count the number of duplicate rows
num_duplicated <- all_trip_data %>%
  dplyr::filter(duplicated(.)) %>%
  nrow()

print(paste("Number of duplicate rows:", num_duplicated))

# Count missing values per column
sum(is.na(all_trip_data))
colSums(is.na(all_trip_data))

# Filter rows with at least one missing value
rows_with_na <- all_trip_data %>% 
  dplyr::filter(if_any(everything(), is.na))

# Display rows with missing values
print(rows_with_na)

# Count the number of rows with missing values
num_rows_with_na <- nrow(rows_with_na)
cat("Number of rows with missing values:", num_rows_with_na, "\n")

# Fill missing values in specific columns
all_trip_data <- all_trip_data %>%
  mutate(
    start_station_name = replace_na(start_station_name, "Unknown Start Station"),
    start_station_id = replace_na(start_station_id, "Unknown Start ID"),
    end_station_name = replace_na(end_station_name, "Unknown End Station"),
    end_station_id = replace_na(end_station_id, "Unknown End ID"),
    end_lat = replace_na(end_lat, median(end_lat, na.rm = TRUE)),
    end_lng = replace_na(end_lng, median(end_lng, na.rm = TRUE))
  )

# Recalculate ride length
all_trip_data <- all_trip_data %>%
  mutate(
    ride_length = as.numeric(difftime(ended_at, started_at, units = "mins"))
  )

# Validate ride length calculations
summary(all_trip_data$ride_length)

# Identify rides with duration <= 0
sum(all_trip_data$ride_length <= 0)

# Remove rows where ride length is 0
all_trip_data <- all_trip_data %>%
  dplyr::filter(ride_length > 0)

# Ensure there are no more invalid ride lengths
sum(all_trip_data$ride_length <= 0)  

# Calculate key metrics: average ride length, total rides by user type
summary_stats <- all_trip_data %>%
  group_by(member_casual) %>%
  summarize(
    avg_ride_length = mean(ride_length, na.rm = TRUE),
    total_rides = n()
  )

# Extract and format the day of the week
all_trip_data <- all_trip_data %>%
  mutate(day_of_week = wday(started_at, label = TRUE, abbr = FALSE))

# Group and analyze trends by user type and day of the week
trends <- all_trip_data %>%
  group_by(member_casual, day_of_week) %>%
  summarize(
    avg_ride_length = mean(ride_length, na.rm = TRUE),
    total_rides = n(),
    .groups = 'drop'
  )

# Sort days of the week in proper order
trends <- trends %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

# Calculate the longest ride duration
max_ride_length <- max(all_trip_data$ride_length, na.rm = TRUE)
print(max_ride_length)

# Identify the most frequent day of the week for rides
mode_day_of_week <- names(which.max(table(all_trip_data$day_of_week)))
print(mode_day_of_week)

# Load additional libraries
library(scales)

# Number of rides per month by user type
rides_per_month <- all_trip_data %>%
  mutate(month = month(started_at, label = TRUE, abbr = FALSE)) %>%
  group_by(member_casual, month) %>%
  summarise(count_rides = n(), .groups = "drop") %>%
  mutate(month = factor(month, levels = month.name))

# Plot number of rides per month
ggplot(rides_per_month, aes(x = month, y = count_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Number of Rides per Month by User Type",
       x = "Month",
       y = "Number of Rides",
       fill = "User Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Number of rides per weekday by user type
rides_per_day <- all_trip_data %>%
  group_by(member_casual, day_of_week) %>%
  summarise(count_rides = n(), .groups = "drop")

# Plot number of rides per weekday
ggplot(rides_per_day, aes(x = day_of_week, y = count_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Number of Rides per Day of the Week by User Type",
       x = "Day of the Week",
       y = "Number of Rides",
       fill = "User Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Calculate average ride duration by user type
ride_duration <- all_trip_data %>%
  group_by(member_casual) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE), .groups = "drop")

# Plot average ride duration
ggplot(ride_duration, aes(x = member_casual, y = avg_ride_length, fill = member_casual)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Average Ride Duration by User Type",
       x = "User Type",
       y = "Average Duration (minutes)",
       fill = "User Type") +
  theme_minimal()

# Peak usage hours analysis
rides_per_hour <- all_trip_data %>%
  mutate(hour = hour(started_at)) %>%
  group_by(member_casual, hour) %>%
  summarise(count_rides = n(), .groups = "drop")

# Plot rides per hour
ggplot(rides_per_hour, aes(x = hour, y = count_rides, color = member_casual, group = member_casual)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = seq(0, 23, by = 2)) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Number of Rides per Hour of the Day by User Type",
       x = "Hour of the Day",
       y = "Number of Rides",
       color = "User Type") +
  theme_minimal()

# Analysis of the most used stations
# Count rides per start station and user type
top_stations <- all_trip_data %>%
  group_by(member_casual, start_station_name) %>%
  summarise(count_rides = n(), .groups = "drop") %>%
  arrange(member_casual, desc(count_rides)) %>%
  group_by(member_casual) %>%
  slice_head(n = 10)  # Select the top 10 most used stations per user type

# Plot Top 10 Most Used Stations by User Type
ggplot(top_stations, aes(x = reorder(start_station_name, count_rides), y = count_rides, fill = member_casual)) +
  geom_col() +
  coord_flip() +  # Flip the chart for better visualization
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Top 10 Most Used Stations by User Type",
       x = "Station Name",
       y = "Number of Rides",
       fill = "User Type") +
  theme_minimal()

```

## Including Plots

```{r, echo=FALSE}

# Plot number of rides per month
ggplot(rides_per_month, aes(x = month, y = count_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Number of Rides per Month by User Type",
       x = "Month",
       y = "Number of Rides",
       fill = "User Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Plot number of rides per weekday
ggplot(rides_per_day, aes(x = day_of_week, y = count_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Number of Rides per Day of the Week by User Type",
       x = "Day of the Week",
       y = "Number of Rides",
       fill = "User Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Plot average ride duration
ggplot(ride_duration, aes(x = member_casual, y = avg_ride_length, fill = member_casual)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Average Ride Duration by User Type",
       x = "User Type",
       y = "Average Duration (minutes)",
       fill = "User Type") +
  theme_minimal()

# Plot rides per hour
ggplot(rides_per_hour, aes(x = hour, y = count_rides, color = member_casual, group = member_casual)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = seq(0, 23, by = 2)) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Number of Rides per Hour of the Day by User Type",
       x = "Hour of the Day",
       y = "Number of Rides",
       color = "User Type") +
  theme_minimal()

# Plot 10 Most used stations by User Type
ggplot(top_stations, aes(x = reorder(start_station_name, count_rides), y = count_rides, fill = member_casual)) +
  geom_col() +
  coord_flip() +  # Flip the chart for better visualization
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("casual" = "#1f78b4", "member" = "#33a02c")) +
  labs(title = "Top 10 Most Used Stations by User Type",
       x = "Station Name",
       y = "Number of Rides",
       fill = "User Type") +
  theme_minimal()

```
